Kindly Bent to Free Us

### 2019-06-06 kindly/motivation.tex
More motivations

\subsection{Hybrid paradigms}

@@

  val get : &('a t) * int -> 'a
  val set : &!('a t) * int * 'a -> unit
end
\end{lstlisting}
    \caption{Affine primitives}
    \label{sig:array}
  \end{subfigure}\hfill
  \begin{subfigure}{0.5\linewidth}
\begin{lstlisting}
let mk_fib_array n =
  let a = create n 1 in
  for i = 2 to n - 1 do
    let x1 = get &a (i-1) + get &a (i-2) in(*@\label{ex:array:get}*)
    set &!a i(*@\label{ex:array:set}*)
  done;
  a(*@\label{ex:array:return}*)
# mk_fib_array : int -> int Array.t
\end{lstlisting}
    \caption{Example of use of affine arrays}
    \label{ex:array}
  \end{subfigure}
\end{figure}

@@

>>>>

\subsection{Imperative programming}

@@

  val get : &('a t) * int -> 'a
  val set : &!('a t) * int * 'a -> unit
end
\end{lstlisting}
    \caption{Affine primitives}
    \label{sig:array}
  \end{subfigure}\hfill
  \begin{subfigure}{0.5\linewidth}
\begin{lstlisting}
let mk_fib_array n =
  let a = create n 1 in
  for i = 2 to n - 1 do
    let x = get &a (i-1) + get &a (i-2) in(*@\label{ex:array:get}*)
    set &!a i x(*@\label{ex:array:set}*)
  done;
  a(*@\label{ex:array:return}*)
# mk_fib_array : int -> int Array.t
\end{lstlisting}
    \caption{Example of use of affine arrays}
    \label{ex:array}
  \end{subfigure}
\end{figure}

Naturally, \lang also checks that borrows do not escape their point
of use through {\em syntactic regions}. The example
above is reproduce in \cref{ex:array:region} with
regions explicitly delimited by \lstinline/{| ... |}/.
\lang ensures that borrows never escape the smallest enclosing
region.
For instance, On \cref{ex:array:region:get}, \lstinline/&a/ does
not escape the region since the borrow is not captured, and the
region only return integers.
Regions annotations are automatically inserted by \lang using
syntactic information such as borrows and binders.
Users can also annotate regions manually to further restrict the range
of a borrow. In practice, we have found that regions
annotations are very rarely needed in common imperative usage.

\begin{figure}
  \centering
  % \begin{subfigure}{0.6\linewidth}
\begin{lstlisting}
let mk_fib_array n =
  let a = create n 1 in
  for i = 2 to n - 1 do {|
    let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
    set &!a i x
  |}
  done;
  a(*@\label{ex:array:return}*)
# mk_fib_array : int -> int Array.t
\end{lstlisting}
    \caption{Example from \cref{ex:array} with explicit regions}
    \label{ex:array:region}
  % \end{subfigure}
\end{figure}

@@

\section{Hybrid programming styles}

While purely functional data-structure present numerous advantages in term
of correctness and ease of use, they sometimes offer worse performances
than their imperative counterpart. On the other hand, reasoning about imperative
data-structure in the context of sharing is notoriously difficult.
%
In this section, we show how linearity and borrowing allows us to define
a safe API for hybrid arrays that support both immutable and mutable features,
and how it can be used to implement a backtracking sudoku solver.

\begin{figure}
  \centering
  \begin{subfigure}{0.48\linewidth}
\begin{lstlisting}
module type HybArray = sig
  type ('a : un) t : aff

  val length : &('a t) -> int
  val get : &('a t) * int -> 'a

  val set : &('a t) * int * 'a -> 'a t
  val set_immut : &!('a t) * int * 'a -> unit
end
\end{lstlisting}
    \caption{Signature}
    \label{sig:hybarray}
  \end{subfigure}\hfill
  \begin{subfigure}{0.5\linewidth}
\begin{lstlisting}
module Cow : HybArray = struct
  include Array

  let set (a, i, x) =
    let f k =
      if k = i then x else Array.get (a,k)
    in
    create (length a) f

  let set_mut = Array.set
end
\end{lstlisting}
    \caption{Implementation using copy-on-write}
    \label{ex:cow}
  \end{subfigure}
  \caption{Hybrid arrays -- signature and example implementation}
\end{figure}

\begin{figure}
  \centering
\begin{lstlisting}[multicols=2]
type board = IntSet.t Mat.t

let propagate_line i0 j0 g n =
  for j = j0+1 to size - 1 do
    let cell = Mat.get &g (i0 , j) in
    let cell' = IntSet.remove n cell in
    Mat.set_mut &!g (i0, j) cell'
  done

let propagate i j g n =
  propagate_line i j &!g n;
  propagate_column i j &!g n;
  propagate_square i j &!g n

let rec solve i j g =
  match is_solved &g with
  | true -> Mat.print &g
  | false ->
    let (new_i, new_j) = next_pos (i,j) in
    let try_solution n =
      let new_g = 
        Mat.set &g (i,j) (singleton n) in
      propagate i j &!new_g n;
      if is_valid &new_g when
        solve new_i new_j new_g
    in
    let cell = Mat.get &g (i,j) in
    IntSet.iter try_solution cell
\end{lstlisting}
  \caption{Chosen excerpt from the sudoku solver}
  \label{ex:cow}
\end{figure}
